### 2015-04-02 zoldatoff/tex/text retrieval/lecture05.tex
Лекции по курсу Text Retrieval
Неделя 2, лекции 1-5

\newpage
\section{Vector Space Retrieval Model: Basic Idea}

@@

\subsection{Vector Space Model (VSM): Illustration}

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{VSM.png}
\end{figure}

\subsection{VSM Is a Framework}
\begin{itemize}
\item Represent a doc/query by a term vector
\begin{itemize}
\item \textbf{Term}: basic concept, e.g., word or phrase
\item Each term defines one dimension
\item N terms define an \textbf{N-dimensional space}
\item \textbf{Query vector}: $q=(x_1, \dots x_N), x_i \in \Re$ is query term weight 
\item \textbf{Doc} vector: $d=(y_1, \dots y_N), y_j \in \Re$ is doc term weight
\end{itemize}
\item $relevance(q,d) \propto similarity(q,d)=f(q,d)$
\end{itemize}

\subsection{What VSM Doesn’t Say}
\begin{itemize}
\item How to define/select the “basic concept” – Concepts are assumed to be orthogonal
\item How to place docs and query in the space (= how to assign term weights)

@@

>>>>

\newpage
\section{Vector Space Retrieval Model}

@@

%----------------------------------------
\subsection{Vector Space Model (VSM): Illustration}

\begin{figure}[H]
    \centering
    \includegraphics[width=0.9\linewidth]{VSM.png}
\end{figure}

%----------------------------------------
\subsection{VSM Is a Framework}
\begin{itemize}
\item Represent a doc/query by a term vector
\begin{itemize}
\item \textbf{Term}: basic concept, e.g., word or phrase
\item Each term defines one dimension
\item N terms define an \textbf{N-dimensional space}
\item \textbf{Query vector}: $q=(x_1, \dots x_N), x_i \in \Re$ is query term weight 
\item \textbf{Doc} vector: $d=(y_1, \dots y_N), y_j \in \Re$ is doc term weight
\end{itemize}
\item $relevance(q,d) \propto similarity(q,d)=f(q,d)$
\end{itemize}

%----------------------------------------
\subsection{What VSM Doesn’t Say}
\begin{itemize}
\item How to define/select the “basic concept” – Concepts are assumed to be orthogonal
\item How to place docs and query in the space (= how to assign term weights)

@@

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{vsm_questions.png}
\end{figure}

%----------------------------------------
\subsection{Simplest VSM = Bit-Vector + Dot-Product + BOW}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{simplest_vsm.png}
\end{figure}

Simplest VSM:
\begin{itemize}
\item Dimension = word
\item Vector = 0-1 bit vector (word presence/absence)
\item Similarity = dot product
\item f(q,d) = number of distinct query words matched in d
\end{itemize}

%----------------------------------------
\subsection{Improved Instantiation}

Improved VSM:
\begin{itemize}
\item Dimension = word
\item Vector = TF-IDF weight vector
\item Similarity = dot product
\end{itemize}

%----------------------------------------
\subsection{Improved VSM with Term Frequency (TF) Weighting}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.6\linewidth]{VSM_TF.png}
\end{figure}

%----------------------------------------
\subsection{IDF Weighting: Penalizing Popular Terms}
IDF — inverse document frequency
\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\linewidth]{IDF.png}
\end{figure}

%----------------------------------------
\subsection{Adding Inverse Document Frequency (IDF)}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.85\linewidth]{VSM_IDF.png}
\end{figure}

%----------------------------------------
\subsection{Ranking Function with TF-IDF Weighting}

\begin{equation*}
f(q, d) = \sum_{i=1}^N x_i \, y_i = \sum_{w \in q \cap d} c(w, q) \: c(w, d) \log \frac{M+1}{df(w)}
\end{equation*}

\begin{itemize}
\item $w \in q \cap d$ - all matched query (q) words in document (d)
\item $c(w, q)$ - count of word w in document d
\item $M$ - total number of documents in collection
\item $df(w)$ - Doc Frequency (total number of documents containing word w)
\end{itemize}

%\begin{figure}[H]
%    \centering
%    \includegraphics[width=\linewidth]{TF_IDF.png}
%\end{figure}

%----------------------------------------
\subsection{TF Transformation: BM25 Transformation}
BM = Best Matching

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{BM25_transformation.png}
\end{figure}

%----------------------------------------
\subsection{Summary}
\begin{itemize}
\item Sublinear TF Transformation is needed to
\begin{itemize}
\item capture the intuition of <<diminishing return>> from higher TF 
\item avoid dominance by one single term over all others
\end{itemize}

\item BM25 Transformation 
\begin{itemize}
\item has an upper bound
\item is robust and effective
\end{itemize}

\item Ranking function with BM25 TF ($k >= 0$):
\end{itemize}

\begin{equation*}
f(q, d) = \sum_{i=1}^N x_i y_i = \sum_{w \in q \cap d} c(w, q) \frac{(k+1) c(w, d)}{c(w, d) + k} \log \frac{M+1}{df(w)}
\end{equation*}

%----------------------------------------
\subsection{Pivoted Length Normalization}

\textbf{Pivoted length normalizer}: use average doc length as <<pivot>>\footnote{Pivot - стержень; точка опоры, вращения}. Normalizer = 1 if $\abs{d}$ = average doc length (avdl).

\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{pivoted_length_norm.png}
\end{figure}

%----------------------------------------
\subsection{State of the Art VSM Ranking Functions}

Pivoted Length Normalization VSM [Singhal et al 96]:
\begin{equation*}
f(q, d) = \sum_{w \in q \cap d} c(w, q) \: \frac{\ln[1+\ln(1+c(w, d))]}{1-b+b\dfrac{|d|}{avdl}} \: \log\frac{M+1}{df(w)}
\end{equation*}

\href{https://ru.wikipedia.org/wiki/Okapi_BM25}{BM25/Okapi} [Robertson \& Walker 94]:
\begin{equation*}
f(q, d) = \sum_{w \in q \cap d} c(w, q) \: \frac{(k+1) \: c(w, d)}{c(w, d) + k\left(1-b+b\dfrac{|d|}{avdl}\right)} \: \log\frac{M+1}{df(w)}
\end{equation*}

%----------------------------------------
\subsection{Further Improvement of VSM?}
\begin{itemize}
\item Improved instantiation of dimension?
\begin{itemize}
\item  stemmed words, stop word removal, phrases, latent semantic indexing (word clusters), character n-grams, ...
\item  bag-of-words with phrases is often sufficient in practice
\item  Language-specific and domain-specific tokenization is important to
ensure “normalization of terms”
\end{itemize}

\item  Improved instantiation of similarity function?
\begin{itemize}
\item  cosine of angle between two vectors?
\item  Euclidean?
\item  dot product seems still the best (sufficiently general especially with appropriate term weighting)
\end{itemize}
\end{itemize}

%----------------------------------------
\subsection{Further Improvement of BM25}
\begin{itemize}
\item BM25F [Robertson \& Zaragoza 09]
\begin{itemize}
\item Use BM25 for documents with structures (<<F>>=fields)
\item Key idea: combine the frequency counts of terms in all fields and then apply BM25 (instead of the other way)
\end{itemize}

\item BM25+ [Lv \& Zhai 11]
\begin{itemize}
\item Address the problem of over penalization of long documents
by BM25 by adding a small constant to TF
\item Empirically and analytically shown to be better than BM25
\end{itemize}
\end{itemize}

%----------------------------------------
\subsection{Summary of Vector Space Model}
\begin{itemize}
\item Relevance(q,d) = similarity(q,d)
\item Query and documents are represented as vectors
\item Heuristic\footnote{Heuristic - эвристический} design of ranking function

\item Major term weighting heuristics 
\begin{itemize}
\item TF weighting and transformation 
\item IDF weighting
\item Document length normalization
\end{itemize}

\item BM25 and Pivoted normalization seem to be most effective
\end{itemize}

\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{pivoted_length_normalization_vsm.png}
\end{figure}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{BM25_Okapi.png}
\end{figure}

%----------------------------------------
\subsection{Recommended reading}
\begin{itemize}
\item A.Singhal, C.Buckley, and M.Mitra. <<Pivoted document length normalization>>. In Proceedings of ACM SIGIR 1996.
\item S. E. Robertson and S. Walker. <<Some simple effective approximations to the 2-Poisson model for probabilistic weighted retrieval>>, Proceedings of ACM SIGIR 1994.
\item S. Robertson and H. Zaragoza. <<The Probabilistic Relevance Framework: BM25 and Beyond>>, Found. Trends Inf. Retr. 3, 4 (April 2009).
\item Y. Lv, C. Zhai, <<Lower-bounding term frequency normalization>>. In Proceedings of ACM CIKM 2011.
\end{itemize}

### 2015-04-02 zoldatoff/tex/text retrieval/lecture06.tex
Лекции по курсу Text Retrieval
Лекции 8 и 9
Лекции по курсу Text Retrieval
Неделя 2, лекции 1-5

\section{Vector Space Retrieval Model: Simplest Instantiation}

\subsection{What VSM Doesn’t Say}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{vsm_questions.png}
\end{figure}

\subsection{Simplest VSM= Bit-Vector + Dot-Product + BOW}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{simplest_vsm.png}
\end{figure}

Simplest VSM:
\begin{itemize}
\item Dimension = word
\item Vector = 0-1 bit vector (word presence/absence)
\item Similarity = dot product
\item f(q,d) = number of distinct query words matched in d
\end{itemize}

>>>>

\section{Implementation of TR Systems}

\subsection{Typical TR System Architecture}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.75\linewidth]{TR_architecture.png}
\end{figure}

%----------------------------------------
\subsection{Tokenization}
\begin{itemize}
\item Normalize lexical units: words with similar meanings should be mapped to the same indexing term
\item Stemming: mapping all inflectional forms of words to the same root form
\item Some languages (e.g., Chinese) pose challenges in word segmentation
\end{itemize}

%----------------------------------------
\subsection{Inverted Index}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\linewidth]{inverted_index.png}
\end{figure}

%----------------------------------------
\subsection{Empirical Distribution of Words}
There are stable language-independent patterns in how people use natural languages:
\begin{itemize}
\item A few words occur very frequently; most occur rarely. E.g., in news articles:
\begin{itemize}
\item Top 4 words: 10~15\% word occurrences 
\item Top 50 words: 35~40\% word occurrences
\end{itemize}
\item The most frequent word in one corpus may be rare in another
\end{itemize}

%----------------------------------------
\subsection{Zipf’s Law}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.95\linewidth]{zipf_law.png}
\end{figure}

\begin{equation*}
F(w) = \frac{C}{r(w)^\alpha}, \alpha \approx 1, C \approx 0.1
\end{equation*}
$\textrm{rank} \times \textrm{frequency} \approx \textrm{constant}$:
\begin{itemize}
\item $F(w)$ - word frequency
\item $r(w)$ - word rank
\end{itemize}

%----------------------------------------
\subsection{Data Structures for Inverted Index}
\begin{itemize}
\item Dictionary: modest size
\begin{itemize}
\item Needs fast random access 
\item Preferred to be in memory 
\item Hash table, B-tree, trie, ...
\end{itemize}

\item Postings: huge
\begin{itemize}
\item Sequential access is expected
\item Can stay on disk
\item May contain docID, term freq., term pos, etc 
\item Compression is desirable
\end{itemize}
\end{itemize}

%----------------------------------------
\subsection{Constructing Inverted Index}

Sort-based method:
\begin{itemize}
\item Step 1: Collect local (termID, docID, freq) tuples from documents
\item Step 2: Sort local tuples by termID (to make <<runs>>) and save to files
\item Step 3: Pair-wise merge runs
\item Step 4: Output inverted file
\end{itemize}

%----------------------------------------
\subsection{Inverted Index Compression}

In general, leverage skewed distribution of values and use variable-length encoding:
\begin{itemize}
\item TF compression:
\begin{itemize}
\item Small numbers tend to occur far more frequently than large
numbers (Zipf's law)
\item Fewer bits for small (high frequency) integers at the cost of more bits for large integers
\end{itemize}

\item Doc ID compression:
\begin{itemize}
\item <<d-gap>> (store difference): $d_1, d_2-d_1, d_3-d_2, \dots$
\item Feasible due to sequential access
\end{itemize}
\end{itemize}

%----------------------------------------
\subsection{Integer Compression Methods}
\begin{itemize}
\item \textbf{Binary}: equal-length coding
\item \textbf{Unary}: $x \geqslant 1$ is coded as $x-1$ one bits followed by 0, e.g.,
3=> 110; 5=>11110
\item \textbf{$\gamma$-code}: x => unary code for $1+\lfloor \log x \rfloor$ followed by uniform code for $x-2^{\lfloor \log x \rfloor}$ in $\lfloor \log x \rfloor$ bits, e.g., 3=>101, 5=>11001
\item \textbf{$\delta$-code}: same as $\gamma$-code, but replace the unary prefix with $\gamma$-code. E.g., 3=>1001, 5=>10101
\end{itemize}

%----------------------------------------
\subsection{General Form of Scoring Function}
\begin{equation*}
f(q,d) = f_a\left( h\left( g(t_1, d, q), \dots, g(t_k, d, q)\right), f_d(d), f_q(q)\right)
\end{equation*}
\begin{itemize}
\item $f_d(d), f_q(q)$ - adjustment factors of document and query
\item $g(t_i, d, q)$ - weight of a \textbf{matched} query term $t_i$ in $d$
\item $h()$ - weights aggregation function 
\item $f_a()$ - final score adjustment function
\end{itemize}

%----------------------------------------
\subsection{A General Algorithm for Ranking Documents}
\begin{itemize}
\item $f_d(d)$ - can be precomputed at index time, $f_q(q)$ - at query time
\item Maintain a score accumulator for each $d$ to compute $h$

\item For each query term $t_i$
\begin{itemize}
\item Fetch the inverted list $\{(d_1,f_1),\dots ,(d_n,f_n)\}$
\item For each entry $(d_j,f_j)$, compute $g(t_i,d_j,q)$, and update score accumulator for doc $d_i$ to incrementally compute $h$
\end{itemize}
\item Adjust the score to compute $f_a$, and sort
\end{itemize}

%----------------------------------------
\subsection{Further Improving Efficiency}
\begin{itemize}
\item Caching (e.g., query results, list of inverted index)
\item Keep only the most promising accumulators
\item Scaling up to the Web-scale? (need parallel processing)
\end{itemize}

%----------------------------------------
\subsection{Some Text Retrieval Toolkits}
\begin{itemize}
\item \href{http://lucene.apache.org/}{Lucene}
\item \href{http://www.lemurproject.org/}{Lemur/Indri}
\item \href{http://terrier.org/}{Terrier}
\item \href{http://meta-toolkit.github.io/meta/}{MeTA}
\item More can be found \href{http://timan.cs.uiuc.edu/resources}{here}
\end{itemize}

%----------------------------------------
\subsection{Summary of System Implementation}
\begin{itemize}
\item Inverted index and its construction 
\begin{itemize}
\item Preprocess data as much as we can 
\item Compression when appropriate
\end{itemize}

\item Fast search using inverted index
\begin{itemize}
\item Exploit inverted index to accumulate scores for documents matching
a query term
\item Exploit Zipf’s law to avoid touching many documents not matching any query term
\item Can support a wide range of ranking algorithms
\end{itemize}

\item Further scaling up using distributed file system, parallel processing, and caching
\end{itemize}

%----------------------------------------
\subsection{Recommended reading}
\begin{itemize}
\item Ian H. Witten, Alistair Moffat, Timothy C. Bell: <<Managing Gigabytes: Compressing and Indexing Documents and Images>>, Second Edition. Morgan Kaufmann, 1999.
\item Stefan B{\"u}ttcher, Charles L. A. Clarke, Gordon V. Cormack: <<Information Retrieval - Implementing and Evaluating Search Engines>>. MIT Press, 2010.
\end{itemize}

### 2015-04-02 zoldatoff/tex/text retrieval/lecture07.tex
Лекции по курсу Text Retrieval
Лекции 8 и 9
Лекции по курсу Text Retrieval
Неделя 2, лекции 1-5

\section{Vector Space Retrieval Model: Improved Instantiation}

Improved VSM:
\begin{itemize}
\item Dimension = word
\item Vector = TF-IDF weight vector
\item Similarity = dot product
\end{itemize}

\subsection{Improved VSM with Term Frequency (TF) Weighting}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{VSM_TF.png}
\end{figure}

\subsection{IDF Weighting: Penalizing Popular Terms}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{IDF.png}
\end{figure}

\subsection{Adding Inverse Document Frequency (IDF)}
\begin{figure}[H]
    \centering
    \includegraphics[width=\linewidth]{VSM_IDF.png}
\end{figure}

>>>>

\newpage
\section{Evaluation of Text Retrieval Systems}

%----------------------------------------
\subsection{The Cranfield Evaluation Methodology}

A methodology for laboratory testing of system components developed in 1960s. General idea is to build reusable test collections and define measures. A test collection can then be reused many times to compare different systems.
\begin{itemize}
\item A sample collection of documents (simulate real document collection)
\item A sample set of queries/topics (simulate user queries)
\item Relevance judgments (ideally made by users who formulated the queries) => Ideal ranked list
\item  Measures to quantify how well a system’s result matches the ideal ranked list
\end{itemize}

%----------------------------------------
\subsection{Evaluating a Set of Retrieved Docs}

\begin{center}
  \begin{tabular}{ | l | c | c | }
    \hline
    & \textbf{Retrieved} & \textbf{Not Retrieved} \\    
    \hline    
    \textbf{Relevant}     & a & b \\ 
    \textbf{Not Relevant} & c & d \\
    \hline  
  \end{tabular}
\end{center}

\begin{itemize}
\item Precision: are the retrieved results all relevant?
\begin{equation*}
Precision = \frac{a}{a+c}
\end{equation*}
\item Recall: have all the relevant documents been retrieved?
\begin{equation*}
Recall = \frac{a}{a+b}
\end{equation*}
\item In reality, high recall tends to be associated with low precision
\end{itemize}

%----------------------------------------
\subsection{Combine Precision and Recall: F-Measure}

\begin{equation*}
F_\beta = \frac{1}{\dfrac{\beta^2}{\beta^2+1}\dfrac{1}{R} + \dfrac{1}{\beta^2+1}\dfrac{1}{P}} = \frac{(\beta^2+1)\cdot P \cdot R}{\beta^2 \cdot P + R}
\end{equation*}

\begin{itemize}
\item $P$ - precision
\item $R$ - recall
\item $\beta$ - parameter, often set to 1: $F_1 = \dfrac{2 \cdot P \cdot R}{P+R}$
\end{itemize}
